---
title: "Flex Dashboard"
author: ""
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: scroll
---
<style type="text/css"> .sidebar { overflow: auto; } </style>

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>

```{r setup, include = FALSE, echo = FALSE}
#rmarkdown::render("lake_ice_flexdashboard.RMD", output_format = "flexdashboard::flex_dashboard")
library(tidyverse)
library(ggplot2)
library(leaflet)
library(htmltools)
library(lubridate)
library(mgcv)
library(gratia)
library(gamm4)
library(itsadug)
library(sp)
library(sf)
library(rgdal)
library(raster)
library(gstat)
library(viridis)


#Working directory for .RMD
knitr::opts_knit$set(echo = TRUE,
                     root.dir = rprojroot::find_rstudio_root_file())

#Controlling figure output in markdown
knitr::opts_chunk$set(
#  fig.height =   
  fig.width = 6,
#  fig.asp = .5,
  out.width = "90%",
#  out.height = 
 fig.align  = "center",
  cache = FALSE,
  eval  = TRUE,
  echo  = TRUE,
  warning = FALSE
)

#Set Theme for ggplot2
theme_set(theme_bw() + theme(plot.title = element_text(hjust = 0.5),
                             plot.subtitle = element_text(hjust = 0.5),
                             legend.position = "bottom"))

#Set Scientific notation output and decimal places for knitr
options(scipen = 999)
options(digits = 4)
options(dplyr.summarise.inform = FALSE)
```

```{r durationclean, include = FALSE}
# Duration Data
duration_spatial <- read.csv("~/advR/ice.trends.github.io/mn/data/lake/duration_spatial.csv")


# Create Leaflet of Average duration
# start with Text Label
duration_map <- duration_spatial %>%
  #all of these func are in {dplyr}
  group_by(DOW, lat, long, pw_basin_n) %>%
  add_count() %>%
  summarise(mean_duration = mean(duration),
            Observations = n) %>%
  distinct(DOW, .keep_all = T) %>%
  mutate(text_label = str_c(
                   pw_basin_n,
                   "<br/>Lake ID: ",
                   DOW,
                   "<br/>Mean Ice Duration: ",
                   round(mean_duration, digits = 1), " Days",
                   "<br/>Number of Observations: ",
                   Observations))
# Pallete for coloring lake points
pal <- colorNumeric(
  palette = "RdBu",
  domain = duration_map$mean_duration)


# creating interpolation map - using sp object

mn_bbox <- readOGR("~/advR/ice.trends.github.io/mn/data/tl_2016_27_cousub.shp") # read in the shape file 

mn_bbox <- spTransform(mn_bbox, CRS("+proj=longlat +ellps=WGS84 +datum=WGS84")) # make sure the data is projected correctly

duration_map_idw <- duration_map %>% ungroup() %>% dplyr::select(lat,long,mean_duration) # keep just the variables of interest

sp_duration <- SpatialPointsDataFrame(cbind(duration_map_idw$long, duration_map_idw$lat), data = duration_map_idw, proj4string = CRS("+proj=longlat +ellps=WGS84 +datum=WGS84")) #create the sp object

sp_duration <- spTransform(sp_duration, CRS("+proj=longlat +ellps=WGS84 +datum=WGS84"))

samp <- spsample(x = mn_bbox, n = 1000000, type = "regular") # get sample plots for the interpolation

duration_idw <- idw(formula = mean_duration ~ 1, locations = sp_duration, newdata = samp, idp = 2) # run interpolation - idp 2 is most common.

raster_idw <- raster(nrow = length(unique(samp@coords[,1])), ncol = length(unique(samp@coords[,2]))) # create raster for the sample points

extent(raster_idw) <- extent(duration_idw) # project to same system

raster_idw <- rasterize(duration_idw, raster_idw, duration_idw$var1.pred) # turn predicted points to cells in raster

duration_idw_leaflet <- leaflet(data = sp_duration) %>%
  addProviderTiles('CartoDB.Positron') %>% 
  addRasterImage(raster_idw, colors = pal(sp_duration$mean_duration), opacity = 0.5) %>% # add on the raster tile
  addCircleMarkers(
    data = duration_map,
    color   = ~pal(mean_duration),
    opacity = 1, #alpha
    weight  = ~3, #outline strength
    radius  = 4 #size of circle
  ) %>%
  addLegend(
    title    = "Ice Duration (Days)",
    position = "bottomright",
    pal = pal,
    values = ~ mean_duration, 
    opacity = 0.9)
```


```{r iceinclean, include = FALSE}
# read Ice-In data
icein_spatial <- read.csv("~/advR/ice.trends.github.io/mn/data/lake/ice_in_spatial.csv")
icein_spatial <- icein_spatial %>%
  mutate(date_in = as.Date(date_in),
         jd_in = ifelse(jd_in < 60, jd_in + 365, jd_in))


# Create Leaflet of Average Ice-on
icein_map <- icein_spatial %>%
  #all of these func are in {dplyr}
  group_by(DOW, lat, long, pw_basin_n) %>%
  add_count() %>%
  summarise(mean_icein = mean(jd_in),
            Observations = n,
            icein_date = as.Date(mean_icein, 
                                origin = as.Date("1960-01-01"))) %>%
  distinct(DOW, .keep_all = T) %>%
  mutate(
    text_label = str_c(pw_basin_n,
                       "<br/>Lake ID: ", DOW,
                       "<br/>Mean Ice-In Date: ",
                       format(icein_date, "%d-%B"),
                       "<br/>Number of Observations: ",
                       Observations))
pal <- colorNumeric(
  palette = "RdBu",
  domain = icein_map$mean_icein, 
  reverse = TRUE)


# creating interpolation map - using sp object


icein_map_idw <- icein_map %>% ungroup() %>% dplyr::select(lat,long,mean_icein)

sp_icein <- SpatialPointsDataFrame(cbind(icein_map_idw$long, icein_map_idw$lat), data = icein_map_idw, proj4string = CRS("+proj=longlat +ellps=WGS84 +datum=WGS84")) 

sp_icein <- spTransform(sp_icein, CRS("+proj=longlat +ellps=WGS84 +datum=WGS84"))

icein_idw <- idw(formula = mean_icein ~ 1, locations = sp_icein, newdata = samp, idp = 2)

raster_idw <- raster(nrow = length(unique(samp@coords[,1])), ncol = length(unique(samp@coords[,2])))

extent(raster_idw) <- extent(icein_idw)

raster_idw <- rasterize(icein_idw, raster_idw, icein_idw$var1.pred)

icein_idw_leaflet <- icein_leaflet <- leaflet(data = sp_icein) %>%
  addProviderTiles('CartoDB.Positron') %>% 
  addRasterImage(raster_idw, colors = pal(sp_icein$mean_icein), opacity = 0.5) %>% 
  addCircleMarkers(
    data = icein_map,
    color   = ~pal(mean_icein),
    label   = ~map(text_label, HTML), #map over labels, make html
    opacity = 0.6, #alpha
    weight  = ~3, #outline strength
    radius  = 4 #size of circle
  ) %>%
  addLegend(
    title    = "Average Ice-On (Julian Days)",
    position = "bottomright",
    pal = pal,
    values = ~ mean_icein, 
    opacity = .9)



```

```{r iceoffclean, include=FALSE}
#read data
iceoff_spatial <- read.csv("~/advR/ice.trends.github.io/mn/data/lake/ice_out_spatial.csv")

iceoff_map <- iceoff_spatial %>%
  #all of these func are in {dplyr}
  group_by(DOW, lat, long, pw_basin_n) %>%
  add_count() %>%
  summarise(mean_iceoff = mean(jd_out),
            Observations = n,
            iceoff_date = as.Date(mean_iceoff, 
                                origin = as.Date("1960-01-01"))) %>%
  distinct(DOW, .keep_all = T) %>%
  mutate(
    text_label = str_c(pw_basin_n,
                       "<br/>Lake ID: ", DOW,
                       "<br/>Mean Ice-Off Date: ",
                       format(iceoff_date, "%d-%B"),
                       "<br/>Number of Observations: ",
                       Observations))
pal <- colorNumeric(
  palette = "RdBu",
  domain = iceoff_map$mean_iceoff)

# idw


iceoff_map_idw <- iceoff_map %>% ungroup() %>% dplyr::select(lat,long,mean_iceoff)

sp_iceoff <- SpatialPointsDataFrame(cbind(iceoff_map_idw$long, iceoff_map_idw$lat), data = iceoff_map_idw, proj4string = CRS("+proj=longlat +ellps=WGS84 +datum=WGS84")) 

sp_iceoff <- spTransform(sp_iceoff, CRS("+proj=longlat +ellps=WGS84 +datum=WGS84"))

iceoff_idw <- idw(formula = mean_iceoff ~ 1, locations = sp_iceoff, newdata = samp, idp = 2)

raster_idw <- raster(nrow = length(unique(samp@coords[,1])), ncol = length(unique(samp@coords[,2])))

extent(raster_idw) <- extent(sp_iceoff)

raster_idw <- rasterize(iceoff_idw, raster_idw, iceoff_idw$var1.pred)

iceoff_idw_leaflet <- leaflet(data = sp_iceoff) %>%
  addProviderTiles('CartoDB.Positron') %>% 
  addRasterImage(raster_idw, colors = pal(sp_iceoff$mean_iceoff), opacity = 0.5) %>% 
  addCircleMarkers(
    data = iceoff_map,
    color   = ~pal(mean_iceoff),
    label   = ~map(text_label, HTML), #map over labels, make html
    opacity = 0.6, #alpha
    weight  = ~3, #outline strength
    radius  = 4 #size of circle
  ) %>%
  addLegend(
    title    = "Average Ice-On (Julian Days)",
    position = "bottomright",
    pal = pal,
    values = ~ mean_iceoff, 
    opacity = .9)

```


### <b><font face="Georgia" size="6em" color="#000000">Ice Duration</font></b>



```{r duration_idw, fig.width=10, fig.height=8, echo=FALSE}
duration_idw_leaflet
```

### <b><font face="Georgia" size="4em" color="#000000">Interpolation of Ice Duration Comments</font></b>
This interactive map is meant to help see the predicted duration of ice for lakes that haven't had data collected for them. These calculations were done using the Inverse Distance Weighting (IDW) method, and done for the entirety of Minnesota. Does your estimate seem accurate? This IDW result seems much more variable than what was expected, with uneven gradients between points. For this reason, I would be wary of placing all of your bets from this map!



### <b><font face="Georgia" size="6em" color="#000000">Ice-On </font></b>

```{r icein_idw, fig.width=10, fig.height=8, echo=FALSE}
icein_idw_leaflet
```

### <b><font face="Georgia" size="4em" color="#000000">Interpolation of Ice-On Comments</font></b>
This interactive map is much like the ice duration interpolation map. It is still more variable than expected, and might not be the best model.

### <b><font face="Georgia" size="6em" color="#000000">Ice-Off</font></b>


```{r iceoff_idw, fig.width=10, fig.height=8, echo=FALSE}
iceoff_idw_leaflet
```